Temporal ordering and registration of images in studies of developmental dynamics

### 2015-02-11 image_analysis_paper_development2/manuscript.tex
added stas edits

Dynamics of developmental progress are commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a broadly applicable manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide software to aid in the application of our methodology to other experimental data sets.
\vspace{1em}

@@

In one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
%
The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \citep{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
%
Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known; typically what is known is
a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
%
Furthermore, images are often collected in different spatial orientations.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
%
%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

Temporal ordering and registration of images can be done manually
when the number of images is small and the differences between them are visually apparent.
%
\fig~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
%
Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
%
In contrast to this example, real data poses nontrivial challenges for both registration and temporal ordering.
%
In general, the landmarks needed for registration, as well as the attributes which can be used to order the data, are not known {\it a priori}.
%
Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.

@@

Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four data sets common in developmental biology.
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
%
Our final data set consists of z-stacks of {\em Drosophila} wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data.
%
We will also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries (such as rotations of the images) which one would like to factor out,
and ``additional" directions of variability (such as temporal dynamics) which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translations and/or scaling via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can be used simply to extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying developmental dynamics.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to confirm our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
%
%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
%
In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images.
%%
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
%
%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps.
%
However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set.
%
Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

From this average trajectory, we can now easily see the developmental progression, which is consistent with the known dynamics.
%
dpERK first appears as two lateral peaks at the ventrolateral side of the embryo; a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventrolateral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks (e.g., fins) and without {\it a priori} knowledge of developmental features (e.g., correlation of age with body size).
%

@@

To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs, which was published in \citet{hamaratoglu2011dpp}.
%
Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs (such as wings, legs, and eyes) in the adult fly during the pupal metamorphosis.
%
The wing disc is an imaginal disc that turns into a wing.
%
The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization.
%
Each disc contains 21 z-slices taken at 1~$\mu$m intervals.
%
The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the molecules that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
%
%The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.

In the wing disc, the anterior-posterior and dorsal-ventral axes are significantly larger than the third principal axis (see \fig~\ref{fig:wing_disc}A). 
%
Therefore, we need not consider registration in all three dimensions, as the z-axis for imaging (the axis perpendicular to the imaging plane) will always be aligned with the short axis of the wing disc.
%
We can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
%
To register the data, we first aligned the maximum intensity projections using angular synchronization.
%
We then used these rotations to register the full three-dimensional data in the x-y plane.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples.
%
Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
%
%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.

We then used diffusion maps to order the registered three-dimensional data.
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
%
In the original data set, each disc was assigned to one of six time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization) by an expert; these times are indicated in \fig~\ref{fig:wing_disc}B.
%
In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
%
The rank correlation coefficient based on the time class is 0.9436.
%
The registration errors are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
%
Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
We can now easily see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
%
%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

%
Vector diffusion maps allow us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with parameters to describe each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several coordinates which concisely and comprehensively describe the data set.
%
This parametrization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
%
Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling, dsilva2013nonlinear}.
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.

>>>>

Dynamics of developmental progress is commonly reconstructed from imaging snapshots of chemical or mechanical processes in fixed embryos.
%
As a first step in these reconstructions, snapshots must be spatially registered and ordered in time.
%
Currently, image registration and ordering is often done manually, requiring a significant amount of expertise with a specific system.
%
However, as the sizes of imaging data sets grow, these tasks become increasingly difficult, especially when the images are noisy and the examined developmental changes are subtle.
%
To address these challenges, we present an automated approach to simultaneously register and temporally order imaging data sets.
%
The approach is based on vector diffusion maps, a manifold learning technique that does not require \textit{a priori} knowledge of image features or a parametric model of the developmental dynamics.
%
We illustrate this approach by registering and ordering data from imaging studies of pattern formation and morphogenesis in three different model systems.
%
We also provide software to aid in the application of our methodology to other experimental data sets.
\vspace{1em}

@@

In one of the common approaches to studies of developmental dynamics, a group of embryos is fixed and stained to visualize a particular biochemical or morphological process within a developing tissue.
%
The developmental dynamics must then be reconstructed from multiple embryos, each of which contributes only a snapshot of the relevant process along its developmental trajectory \citep{jaeger2004dynamic, peter2011gene, fowlkes2008quantitative}.
%
Importantly, the ``age'' of any given embryo arrested in its development is often only approximately known.
%
Typically, what is known is
a certain time window to which a collection of embryos belongs \citep{ng2012large, richardson2014emage, castro2009automatic}.
%
Furthermore, images are often collected in different spatial orientations.
%
In order to recover the developmental dynamics from such data sets, snapshots of different embryos must first be spatially aligned or {\em registered}, and then ordered in time.
%
%We show how recently developed dimensionality reduction algorithms can automate {\it and combine} both of these tasks.

Temporal ordering and registration of images can be done manually
when the number of images is small and the differences between them are visually apparent.
%
\fig~\ref{fig:fish} shows a caricature of fish development which illustrates the processes of growth and patterning.
%
In this case, temporal ordering can be accomplished by arranging the fish by size, which is monotonic with the developmental progress.
%
Image registration is based on obvious morphological landmarks, such as the positions of the head and the fins.
%
In contrast to this example, real data pose nontrivial challenges for both registration and temporal ordering.
%
In general, the landmarks needed for registration, as well as the attributes which can be used to order the data, are not known {\it a priori}.
%
Additional challenges arise from embryo-to-embryo variability, sample size, and measurement noise.

@@

Here, the vector diffusion maps algorithm is adapted for the analysis of images of developing tissues in studies of developmental dynamics, with the main objective of revealing stereotypic developmental trajectories from fixed images.
%
To illustrate our approach, we analyze four experimental data sets.
%
Our first two data sets come from live imaging studies of {\em Drosophila} and zebrafish embryogenesis.
%
In both of these examples, the correct rotational orientation and temporal order are independently known, and these data sets will be used to validate our approach.
%from a study of {\it Drosophila} embryogenesis, one of the best experimental models for studies of development \citep{jaeger2012drosophila}.
%
Our third data set consists of images from fixed {\em Drosophila} embryos where the correct orientation and order is unknown; here, we will show how the algorithm can help uncover developmental dynamics which are not readily apparent.
%
Our final data set consists of z-stacks of {\em Drosophila} wing discs, which we will use to illustrate how our methods can be used to analyze specific types of three-dimensional imaging data.
%
We also show how to compute an average trajectory from a set of registered and ordered fixed images to remove noise due to intersample variability and obtain a smooth description of the underlying developmental dynamics.

@@

Vector diffusion maps \citep{singer2012vector} is a manifold learning
technique developed for data sets which contain two sources of variability:
geometric symmetries, such as rotations of the images, which one would like to factor out,
and ``additional" directions of variability, such as temporal dynamics, which one would like to uncover.
%
Vector diffusion maps combine two algorithms, {\em angular synchronization} \citep{singer2011angular} for image registration and {\em diffusion maps} \citep{coifman2005geometric} for extracting intrinsic low-dimensional structure in data, into a single computation.
%
We will use the algorithm to register images of developing tissues with respect to planar rotations, as well as uncover the main direction of variability {\it after} removing rotational symmetries.
%
Although in general, images may contain variations due to rotations, translations, and scaling, we will remove the relevant translations and/or scaling via relatively simple image preprocessing, and focus only on factoring out rotations using the vector diffusion maps algorithm.
%
In the case that {\em all} relevant symmetries can be removed with straightforward preprocessing, our algorithms can extract the main direction of variability within the imaging data set.
%
We assume that this main direction of remaining variability in these images is parameterized by the developmental time of each embryo.
%
As a consequence, uncovering this direction should reveal the underlying dynamics.

@@

After removing variability due to rotations, the developmental dynamics may be revealed by ordering the data along the one-dimensional curve that parameterizes most of the remaining variability in the data.
%
Such a curve can be discovered using diffusion maps \citep{coifman2005geometric}, a nonlinear dimensionality reduction technique that reveals a parametrization of data that lies on a low-dimensional manifold in high-dimensional space.
%
The idea is illustrated in \fig~\ref{fig:schematics}C, where the data are two-dimensional points which lie on a one-dimensional (nonlinear) curve.
%
We use {\it local} information about the data to find a parametrization which respects the underlying manifold geometry, so that points which are close in high-dimensional space (e.g., images which look similar) are close in our parametrization.
%
This idea of locality is denoted by the color of the edges in \fig~\ref{fig:schematics}C:
data points which are close are connected by dark edges, and clearly, the dark edges are more ``informative" about the low-dimensional structure of the data.
%
The color in \fig~\ref{fig:schematics}D depicts the one-dimensional parametrization or ordering of the data that we can detect visually.
%
A detailed example of using vector diffusion maps to register and order synthetic data is given in \fig~\ref{fig:1d_example}, and a step-by-step tutorial of the diffusion maps implementation is included in the \SI.
%
In our working examples, each data point will be of much higher dimension (e.g., a pixelated image or three-dimensional voxel data), and so we cannot extract this low-dimensional structure visually.
%
Instead, we will use diffusion maps which automatically uncovers a parametrization of our high-dimensional data from the eigenvectors of the appropriate matrix (see \SI).
%
Furthermore, the corresponding eigenvalues will allow us to test our assumption that our data approximately lie on a one-dimensional manifold (see \fig~\ref{fig:eigenvalues_drosophila_live}--\ref{fig:eigenvalues_wing_disc}).
%
%Diffusion maps generalizes directly from one-dimensional nonlinear curves to higher-dimensional manifolds.

@@

To assess the robustness of the proposed methodology, we repeated this procedure with four additional data sets extracted from independent live imaging studies spanning the same developmental time period.
%
The results are shown in \fig~\ref{fig:drosophila_live_imaging}E.
%
The errors in the recovered angles are all less than 10$^\circ$, and the rank correlation coefficients are consistently greater than 95\%, indicating that our methodology can reproducibly order data of this type.
%
%In general, to accurately recover the temporal order, the vector diffusion maps algorithm requires a sufficient amount of data so that the underlying one-dimensional curve (i.e., the curve in \fig~\ref{fig:schematics}C) is well-sampled.
%
%At the same time, the accuracy of the recovered alignments is a function of the signal-to-noise ratio in the pairwise alignments of the images, as well as the size of the data set \citep{singer2011angular}.
%
%Clearly, $40$ images are sufficient to accurately recover the rotational orientation and temporal order in this specific developmental time frame.
%%
%To quantitatively assess how our methods perform as a function of sample size, we used bootstrap sampling to compute the rank correlation as a function of the size of the data set.
%%
%Subsets of images were randomly selected from the $40$-image data set, and then registered and ordered.
%%
%\fig~\ref{subfig:movie_bootstrap} shows the median rank correlation coefficient as a function of the number of images in the subset.
%%
%As expected, the accuracy in the recovered ordering increases with the number of images.
%%
%At the same time, we see that correct order is recovered even with a relatively small number of snapshots, which should make our approach applicable other studies of developmental processes.
%%
%Because our particular data set contains relatively low noise levels, few images were required to obtain accurate rotational alignment, and the high temporal resolution of our live imaging study allows us to order the images accurately.
%
%In summary, the algorithm accurately and robustly registers and orders a model data set where both the rotation angles and real times are independently known.

@@

\fig~\ref{fig:drosophila_fixed_images}B shows the selected images in \fig~\ref{fig:drosophila_fixed_images}A, now registered and ordered using vector diffusion maps.
%
Registered and ordered images of individual embryos can then be used to construct a representative average trajectory. 
%
%However, because of interembryo variability, these selected images are not entirely representative of the developmental dynamics across the entire data set.
%
%To highlight the developmental dynamics revealed by vector diffusion maps, we computed an average developmental trajectory across the set of registered and ordered images (see \SI).
%
Each snapshot in the average trajectory is the (weighted) average of a group of successive images from the registered and ordered data set (see \SI).
%
Averaging successive images removes some of the interembryo variability, so that sequential snapshots of this averaged trajectory, shown in \fig~\ref{fig:drosophila_fixed_images}C, serve as a summary of the stereotypic developmental dynamics.

@@

From this average trajectory, we can now easily see the developmental progression consistent with the known dynamics:
%
dpERK first appears as two lateral peaks at the ventrolateral side of the embryo, and a third dpERK peak then appears at the dorsal side of the embryo.
%
During mesoderm invagination, the two ventrolateral dpERK peaks merge together, eventually forming, together with Twi, the ``omega'' shape.
%
The dorsal dpERK peak then disappears during germband extension as cells from the ventral side wrap around to the dorsal side.
%
At the end of this process, similar ``omegas'' formed by Twi and dpERK appear on the dorsal side of the embryo; these patterns are most readily seen in the last image of \fig~\ref{fig:drosophila_fixed_images}C.
%
Thus, vector diffusion maps can accomplish the tasks presented in the caricature in \fig~\ref{fig:fish}, even in the absence of information about image landmarks and without {\it a priori} knowledge of developmental features.
%

@@

To demonstrate this approach, we used an existing three-dimensional data set of fixed {\em Drosophila} wing imaginal discs \citep{hamaratoglu2011dpp}.
%
Imaginal discs are groups of progenitor cells in fly larva that will transform into specific organs during metamorphosis.
%
The wing disc is an imaginal disc that turns into a wing.
%
The data set is composed of 46 fixed wing discs whose developmental times range from 72 to 112~hours after fertilization.
%
Each disc contains 21 z-slices taken at 1~$\mu$m intervals.
%
The discs were dissected from larvae expressing the Dad-GFP reporter construct (green) and stained with antibodies that recognize Spalt (red), Wingless (gray), and Patched (gray), the factors that play important roles in disc patterning and growth (\fig~\ref{fig:wing_disc}A).
%
%The disc grows in time; at long times, the disc becomes too large to fit completely within the frame and only a partial disc is imaged.

In the wing disc, the anterior-posterior and dorsal-ventral axes are significantly larger than the third principal axis (see \fig~\ref{fig:wing_disc}A). 
%
Therefore, we need not consider registration in all three dimensions, and can instead focus on registering the wing discs with respect to rotations only in the x-y plane.
%
To register the data, we first aligned the maximum intensity projections using angular synchronization.
%
We then used these rotations to register the full three-dimensional data in the x-y plane.
%
Because the maximum intensity projections are two-dimensional images, this step is no more computationally intensive than the previous examples.
%
Such an approach is possible when there are distinct major and minor axes within a three-dimensional sample, which reduces the rotational degrees of freedom.
%
%Other options for registering samples include using a specific localized signal (such as Twi in \fig~\ref{fig:drosophila_fixed_images}) as a marker for alignment, or using other {\em a priori} knowledge.

We then used diffusion maps to order the registered three-dimensional data.
%
\fig~\ref{fig:wing_disc}B shows selected images from the data set ordered by diffusion maps.
%
In the original data set, each disc was assigned to one of six time classes (72--73~hr, 76.5--77.5~hr, 79--80~hr, 89--90~hr, 100--101~hr, and 110.5--111.5~hr after fertilization) by an expert; these times are indicated in \fig~\ref{fig:wing_disc}B.
%
In the ordered set, the size of wing disc grows, and the intensity of the Dad-GFP signal increases as a function of time.
%
The rank correlation coefficient based on the time class is 0.9436.
%
The registration errors are primarily due to some wing discs having extra tissue attached to them (such as the image in \fig~\ref{fig:wing_disc}A and the fourth image in \fig~\ref{fig:wing_disc}B).
%
Even with such obstructions, we can accurately order the images and extract a stereotypical developmental trajectory, shown in \fig~\ref{fig:wing_disc}C, by averaging (see \SI).
%
We can now clearly see the growth of the wing disc, even though averaging somewhat blurs some finer scale structures.
%
%Thus, this result shows that the algorithm can successfully order a three-dimensional data set.

@@

%
Vector diffusion maps allow us to automatically register images, an essential task for many applications.
%
Simultaneously, the algorithm provides us with parameters to describe each image.
%
In the examples presented here, we have focused on ordering the images in time using the first vector diffusion maps coordinate.
%
In general, we can recover several coordinates which concisely and comprehensively describe the data set.
%
This parametrization can then be used for typical data analysis tasks, such as outlier detection and model fitting.
%
%Data which comes from several different genetic backgrounds can be clustered according to their spatial expression patterns, and clustering can also be used to separate images into different developmental stages.
%
Furthermore, images taken from different viewing directions can be analyzed, as the vector diffusion maps parametrization will organize the images according to the viewing angle \citep{singer2011viewing}.
%
Another direction for future work is related to the joint analysis of data sets provided by different imaging approaches, such as merging live imaging data of tissue morphogenesis with snapshots of cell signaling and gene expression from fixed embryos \citep{krzic2012multiview, ichikawa2014live, rubel2010coupling, dsilva2013nonlinear}.
%
In the future, it would also be interesting to explore the connections between our proposed approach and recently developed methods for ordering and classification of face images \citep{kemelmacher2011exploring, kemelmacher2014illumination}
%
Given the rapidly increasing volumes of imaging data from studies of multiple developmental systems, we expect that dimensionality reduction approaches discussed in this work will be increasingly useful for biologists and motivate future applications and algorithmic advances.
